Synthesising Recursive Functions for\\ First-Order Model Counting

### 2023-04-11 doc/talks/3_long/talk.tex
First version of the summary part of the talk

\author{Paulius Dilkas \\ \vspace{1cm} \tiny{Joint work with Vaishak Belle (Univeristy of Edinburgh, UK)}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{National University of Singapore, Singapore}
\date{18th April 2023}
% TODO: unify the use of colours, maybe replace red!50 with one of the main colours
% TODO: how do we capitalise titles of \block?
% TODO: make sure I'm using the 'forest' package whenever possible

@@

\begin{frame}{Back to Our Example}
  The following function counts injections:
  \[
  f(m, n) =
  \begin{cases}
    1 & \text{if } m = 0 \text{ and } n = 0 \\
    0 & \text{if } m > 0 \text{ and } n = 0 \\
    f(m, n-1) + mf(m-1, n-1) & \text{otherwise.}
  \end{cases}
  \]
  \pause
  \begin{itemize}
    \item \alert{$f(m, n)$} can be computed in \alert{$\Theta(mn)$} time
    \begin{itemize}
      \item using dynamic programming
    \end{itemize}\pause
    \item \alert{Optimal} time complexity to compute \alert{$n^{\underline{m}}$}
          is \alert{$\Theta(m)$}
    \item But \alert{$\Theta(mn)$} is still much better than translating to
          propositional logic and solving a $\#\P$-complete problem \pause
    \item The rest of this talk is about how to find such functions automatically
  \end{itemize}
\end{frame}

@@

% TODO: need 3 slides in total for the ending

% really focus on the 'so what?'
\begin{frame}{Summary \& Future Work}
  \begin{block}{Summary}
    The circuits hitherto used for FOMC become more powerful with:
    \begin{itemize}
    \item cycles,
    \item generalised domain recursion,
    \item and some more new compilation rules that support domain recursion.
    \end{itemize}
  \end{block}
  \begin{block}{Future Work}
    \begin{itemize}
    \item Automate:
      \begin{itemize}
      \item extracting and simplifying the definitions of functions,
      \item finding all base cases.
      \end{itemize}
    \item Open questions:
      \begin{itemize}
      \item What kind of \alert{sequences} are computable in this way?
      \item Would using a \alert{different logic} extend the capabilities of FOMC further?
      \end{itemize}
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}{Major Directions for Future Work}
  \begin{itemize}
    \item An algebraic description of what kind of sequences and functions with
          domain $\mathbb{N}_{0}^{k}$ are computable in this way
          \begin{itemize}
            \item monotonicity, maximal growth rate, etc.
          \end{itemize}
    \item A different input format or logic that allows the same approach to
          capture more computations
          \begin{itemize} % TODO: come back to the 'many ways to count' slide
                          % and add more arrows (e.g., beyond FOL)
            \item fixed-point logic with counting
            \item `let domain $\Delta \coloneqq \{\, 1, 2, \dots, MC(\phi) \,\}$'
          \end{itemize}
    \item parameterised Markov logic networks
    \begin{itemize}
      \item `What equations do the domain sizes have to satisfy for the
            probability of event $E$ to be at least $95\%$?'
    \end{itemize}
  \end{itemize}
\end{frame}

>>>>

\author{Paulius Dilkas \\ \vspace{1cm} \tiny{Joint work with Vaishak Belle (Univeristy of Edinburgh, UK)}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{National University of Singapore, Singapore}
\date{18th April 2023}
% TODO: unify the use of colours, maybe replace red!50 with one of the main colours
% TODO: how do we capitalise titles of \block?
% TODO: make sure I'm using the 'forest' package whenever possible
% TODO: consider adding an outline slide
% TODO: add periods at the end of sentences

@@

\begin{frame}{Back to Our Example}
  The following function counts injections:
  \[
  f(m, n) =
  \begin{cases}
    1 & \text{if } m = 0 \text{ and } n = 0 \\
    0 & \text{if } m > 0 \text{ and } n = 0 \\
    f(m, n-1) + mf(m-1, n-1) & \text{otherwise.}
  \end{cases}
  \]
  \pause
  \begin{itemize}
    \item \alert{$f(m, n)$} can be computed in \alert{$\Theta(mn)$} time
    \begin{itemize}
      \item using dynamic programming
    \end{itemize}\pause
    \item \alert{Optimal} time complexity to compute \alert{$n^{\underline{m}}$}
          is \alert{$\Theta(m)$}
    \item But \alert{$\Theta(mn)$} is still much better than translating to
          propositional logic and solving a \alert{$\#\P$-complete} problem
          \pause
    \item The rest of this talk is about how to find such functions automatically
  \end{itemize}
\end{frame}

@@

\begin{frame}{What Have We Learned?}
  \begin{itemize}
    \item Knowledge compilation can build \alert{graphs with cycles}.
    \item Graphs (as well as circuits) define \alert{functions}.
    \item Cycles can represent \alert{recursive calls}, including:
    \begin{itemize}
      \item mutual recursion
      \item and function calls as complex as \alert{$f(n-k-2)$}.
    \end{itemize}
    \item Recursion helps us solve counting problems that were \alert{previously
          beyond the reach of FOMC}\@.
    \item In some cases, even if a polynomial-time solution is already known,
          Crane is able to find \alert{more efficient solutions}, with a lower
          degree polynomial.
  \end{itemize}
\end{frame}

% TODO: maybe add a future <-> past picture slide

\begin{frame}[fragile]{Beyond First-Order Logic}
  \vspace{0.5cm}
  \hspace*{0.25\linewidth}
  \begin{tikzcd}[remember picture]
    |[alias=WMI]|\WMI{} & \WFOMI{} & \\
    |[alias=WMC]|\#\SAT{}/\WMC{} \arrow[u] \ar[d] \ar[r] & \textup{(W)FOMC} \arrow[u] \ar[d] \ar[r,dashed] & \raisebox{-0.4\height}{\includegraphics[width=0.1\linewidth]{question-mark.jpg}} \\
    |[alias=PBP]|\AMC{}, \pbp{} & |[alias=SP]|\SPPP{} &
  \end{tikzcd}
  \begin{tikzpicture}[overlay,remember picture]
    \begin{scope}[transform canvas={xshift=-3em}]
      \draw[->,draw=gray] ($(WMC)+(0,0.1)$)--($(WMI)+(0,0.5)$) node[midway,left] {\textcolor{gray}{$\sum \to \int$}};
      \draw[->,draw=gray] ($(WMC)-(0,0.1)$)--($(PBP)-(0,0.5)$) node[midway,left] {\textcolor{gray}{Generalising weights}};
    \end{scope}
    \begin{scope}[transform canvas={yshift=-2em}]
      \draw[->,draw=gray] ($(PBP)-(1,0)$)--($(SP)+(1,0)$) node[midway,below] {\textcolor{gray}{From propositional to first-order logic}};
    \end{scope}
  \end{tikzpicture}
  \vfill
  \vspace{1.5cm}
  What kind of logic is needed to succinctly describe, e.g.,
  \begin{itemize}
    \item $f(n) = f(f(n-1))$
    \item or the Fibonacci sequence?
  \end{itemize}
\end{frame}
% impossibility results as well (finite model theory)

\begin{frame}{Algebraic Solutions to Parameterised Problems (1/2)}
  \begin{itemize}
    \item Suppose we have a Markov logic network that models the probability
          \alert{$P$} that some system will fail.
    \item Here:
    \begin{itemize}
      \item domain sizes describe the numbers of various components,
      \item and weights express probabilities that:
      \begin{itemize}
        \item some component fails,
        \item or some combination of failures leads to another failure.
      \end{itemize}
    \end{itemize}
    \item Crane can express \alert{$P$} as a function of the domain sizes and
          weights.
  \end{itemize}
\end{frame}

\begin{frame}{Algebraic Solutions to Parameterised Problems (2/2)}
  With the help of a computer algebra system, we can then:
  \begin{itemize}
    \item determine how \alert{$P$} scales with the number of users,
    \item find combinations of domain sizes that keep \alert{$P$} below some
          threshold,
    \item find ranges of weights that keep \alert{$P$} sufficiently small across
          a range of domain size values.
  \end{itemize}
\end{frame}
% TODO: add a concluding diagram: 'reasoning with functions'(in a box) >(big) 'reasoning with numbers'
